---
title: "Lab 1"
author: "IMC 490: Machine Learning for IMC"
output: pdf_document
---

In this lab, we'll be going over the following topics:

- operations and data structures
- dataframes
  - accessing data
  - indexing data
  - data types
  - gotchas
- regression

Website for Introduction to Statistical Learning:
http://www-bcf.usc.edu/~gareth/ISL/

**Remember to use ?command or help(command) in the R console to access documentation at any time if you have questions.**

### Navigation

`setwd()`  
`getwd()`  
`list.files()`  

The above three commands will allow you to navigate your filesystem. You'll use `getwd()` (**get** **w**orking **d**irectory) to figure out where you are, `setwd()` (**set** **w**orking **d**irectory) to move around, and `list.files()` to look at the stuff that's in the folder.
 

```{r,}
setwd("~/Documents/Machine-Learning-IMC490/Lab1/")
```

```{r}
getwd()
```

```{r}
list.files()
```

**Common gotcha:** Be sure to enter the filepath in `setwd()` as a string (in quotes " "). A common mistake is to forget the quotes. If you do, R will think that */your/filepath/* is a variable holding some value... And complain when it finds that it isn't.

\newpage
### Data Types and Vectors
`c()`  

From the R documentation: "R has six basic (‘atomic’) vector types: logical, integer, real, complex, string (or character) and raw."  

We often use integer, real, string, and logical types. In normal use we almost never see complex and raw.

Let's make some vectors. `c()` (**c**ombine) is the generic function for creating a vector.  

```{r}
# 2 ways of making a vector with numbers 1 through 5
c(1, 2, 3, 4, 5)
```

```{r}
c(1:5)
```

```{r}
# create a character (string) vector and check its type
a_char_vec = c("a", "b", "c")
typeof(a_char_vec)
```

```{r}
# create a boolean vector
some_numbers = c(1:5)
a_bool_vec = some_numbers >= 3
a_bool_vec
```

The operation *>= 3* was vectorized and applied to each element of some_numbers. Vectorized operations are at the core of R. A vectorized operation is an operation that is applied to each element of a vector. This includes arithmetic and comparison operations.

```{r}
# create a vector 1:10 and add 5 to each element
x = c(1:10)
x + 5
```

**Common gotcha:** If you perform an operation with two vectors of different length, the shorter vector will be extended to complete the operation.

```{r}
x = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
y = c(0, 100)
x + y
```

\newpage

**Some more vector functions:**

Casting vectors from one type to another:  
`as.integer()`  
`as.numeric()`  
`as.character()`  
  
Checking characteristics of a vector:
`typeof()`    
`length()`  
`unique()`  
  
Calculating statistics of a numeric vector:  
`mean()`    
`sd()`  
`min()`  
`max()`  

\newpage
### Dataframes
`data.frame()`  

A dataframe is simply a collection of vectors. Here we're making a dataframe with column x which is an index, and y, which are random samples from a standard normal distribution.  

```{r}
dat = data.frame(x = c(1:10), y = rnorm(10))
dat
```

To extract vectors from a dataframe, use the dollar sign operator.

```{r}
dat$y
```


#### Reading in data
`read.csv()`  
`data()`

Link to sample dataset: http://www-bcf.usc.edu/~gareth/ISL/Auto.csv

To read data into R, use the `read.csv()` function with a filepath or a URL.  

```{r}
read.csv("")
```








